Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy

### 2021-07-28 cover_letter/response_letter_MJ_210721.tex

\begin{figure}[ht!]
	\centering
    \includegraphics[angle=90,origin=c,width=1.0\textwidth]{FigS7.pdf}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \caption*{\textbf{Figure S7.}\rood{Macrostate interpretation and structural analysis. The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble. (a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying. We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction. This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction. We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over the two central bases, represented by red arrows, in the complementary strands. It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands. We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift. We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns. (b) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence. The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate. Cartoon renderings reflect the dominant configurations of each macrostate and how each is represented by the $d_{3^\prime}^{-1}$ and $d_{5^\prime}^{-1}$ order parameters green and red arrows, respectively. (c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$. This panel conveys two important pieces of information. First. each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble. These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail. Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex. Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$. A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and a an increase in $d_{5^\prime}^{-1}$ (increase in $d_{5^\prime}$) producing the stable 3S2 state (green). A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black). Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple). The infinitely separated dehybridized state D exists at the origin, but limited by the finite box size corresponds to $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red). The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends. (d) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments. The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another.}}

@@

\textcolor{blue}{4)      In Figure 1 the units for rates are not provided.}

\textbf{Author Reply:}   Thank you for catching this error. Fig.~1 has been updated to include units in s$^{-1}$.

\textcolor{blue}{5)      In the comparison between experiments and theory for slow and fast T-jump IR responses (P. 17 L. 9), were there any correlations between the goodness-of-fit in the calculated rates (Figure S4) and the agreement with experiments for a given system? It is hard to judge as is because the goodness-of-fit values were not provided with Figure S4.}

\textbf{Author Reply:}     We have updated all panels of Fig.~S4 plots to include $R^2$ goodness-of-fit values determined by linear least squares fits of the models in log-space (i.e., $\log \left( f \right) = -k_d t$) and have also re-plotted the data on log-linear axes for easier comparisons between the data and fits. The revised figure is reproduced below. In all cases except the lowest temperature (T$_m$-5) slow response we report good fits with $R^2$ $>$ 0.88. We see marginally less good fits for the GC-core sequence compared to the other three systems, possibly -- as we observe in Section 3.1 of the main text -- due to its high propensity to fray, but the reduction in the quality of the fit is only marginal and in absolute terms the fits are still very good.

@@

``A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of \rood{7 mM}. \rood{This concentration is only 3.5$\times$ larger than the 2 mM concentration employed in our experimental analyses (Section 2.2).}''

(III) Finally, we deliberately chose to explore the melting temperature as the state point at which the hybridized and dehybridized populations are in balance and sought to probe the mechanism and kinetics of the hybridization/dehybridization transitions under these equilibrium conditions corresponding to no net thermodynamic preference for either state (i.e., minimally driven conditions). A favorable side effect of this choice was indeed that we observe approximately equal numbers of hybridization and dehybridization events in a single (sufficiently long) simulation, whereas temperatures away from $T_\mathrm{melt}$ would favor one over the other. It is an interesting question to consider how the mechanism and kinetics of hybridization/dehybridization may change with temperature. Exploring small excursions either side of $T_\mathrm{melt}$ may still be possible using unbiased simulations, but we concur with the reviewer that biased sampling would be required to observe sufficient numbers of hybridization (dehybridization) events at large excursions above (below) $T_\mathrm{melt}$.

In response to this point, we have modified text in the terminal paragraph of the Conclusions on p.~38 observing that this may be an interesting point for follow-up study:

``Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons. \rood{We also suggest that the MSM approach followed in this work, possible coupled with biased sampling and reweighting techniques, \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels} may be well-suited to exposing changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.}''

@@

(III) We greatly appreciate this comment and suggestion to apply a fluctuation correction to our data in order to correct for artifacts associated with finite-size effects following the approach detailed in Ref.~\citep{Ouldridge2010ExtractingSimulations}. We have reanalyzed our slow repsonse data under the homodimer correction and this has pleasingly resulted in a slightly improved agreement between our simulations and experiment. The revised figure is reproduced below.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Fig1.pdf}
        \caption*{\textbf{Figure 1.} Experimental measurements and computational predictions of slow and fast at T-jump IR responses. Results are reported in terms of the final T-jump temperature. (a) The experimental and simulated slow rate constants $k_d^\mathrm{slow}$ corresponding to duplex dissociation over long time scales. \rood{We apply a fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to account for finite-size effects and ensure an asymptotically correct dimer fraction} (b) The experimental and simulated fast rate constant $k_d^\mathrm{fast}$ corresponding to terminal base-pair fraying on short time scales. The simulation results are corrected by a sequence-independent scaling factor that corrects for a 10$\times$ acceleration of the slow dissociation dynamics and 120$\times$ acceleration of the fast fraying dynamics. The simulated temperature in all cases is subjected to a (+4) K corrective calibration to account for an observed systematic under-prediction of the melting temperature by the 3SPN.2 model}
        \label{fig:relaxation-comparison}
	\end{center}
\end{figure}

In light of this modification, we have also added the following text to the Section 3.1 on p.~16:

\rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box and ensure that we will observe asymptotically correct behavior at long timescales, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to our predictions for $f_\mathrm{hybridized}$ bulk yields. We correct the values observed in our simulations $f_\mathrm{hybridized}^\mathrm{sim}$ to the bulk corrected result $f_\mathrm{hybridized}$ using the following relations:}

    \begin{equation}\label{}
    \phi = \frac{f_\mathrm{hybridized}^\mathrm{sim}}{1 - f_\mathrm{hybridized}^\mathrm{sim}}
    \end{equation}

@@

``Going forward, we will extend this work to discern more general trends in sequence-dependent hybridization/dehybridization for a wider range of oligomer sequences and motivate strategies for experimental comparisons. \rood{We also suggest that the MSM approach followed in this work, possibly coupled with biased sampling and reweighting techniques \citep{Wu2014StatisticallyStates, Scherer2015PyEMMAModels}, may be well-suited to exposing changes in the hybridization/dehybridization mechanism and kinetics as a function of temperature.}''

\textcolor{blue}{7. The authors could explain a bit more carefully (perhaps in the SI) how they handle the multiple nearest-neighbour states that correspond to their microstates (eg. F4 clearly involves more than one NN configuration).}

We have modified the text in the ``Nearest neighbor model of duplex thermodynamics'' text in the Supporting Information to better describe for the reader how we performed this calculation. Specifically, we have appended the following text at the end of this discussion:

``\rood{In applying the NN models to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a. Indeed, each macrostate is composed of an ensemble of partially frayed states with higher free energy contributions than the intact pattern. However, when averaging across the frayed contributions of all microstates in each macrostate, we have found that incorporating additional NN configurations changes 
$\Delta\Delta$ F values by $<$ 1 kJ/mol when compared to using a single microstate. Given the extent to which a single NN representation captures each macrostate (which is further demonstrated in Fig.~S7), we have elected to maintain a single microstate free energy estimate for each macrostate}''

\noteg{Can add some more text here include a figure for the F4 or shifted states showing a minimal difference.}

@@

\textcolor{blue}{8. I don't understand how the population of F4 could be >50\% that of H for the GC-core system, when the free energy of F4 is reported as > 5kJ/mol. My back-of-the-envelope maths predicts a ratio of 8:1 for a 5kJ/mol gap. Incidentally, 8:1 is roughly the ratio I got from running the configurations through NUPACK.}

\textbf{Author Reply:}  We apologize for the confusion and have found that it stems from a unit conversion error for the GC-core sequence. Specifically, we did not convert the $\Delta\Delta$ F (5.1 kT) from units kT to kJ/mol, which should result in a much larger $\Delta\Delta$ F (13.7 kJ/mol) relative to the hybridized state. 
Using this correct $\Delta\Delta$ F value we see agreement between the reported free energy in Fig~3 and populations reported in Fig.~2. In brief, we perform the following back-of-the-envelope verification:

\begin{itemize}
\item $G_H$ = $G_D$ = 0 (adopting this as the arbitrary zero of free energy and asserting that the sequence is at its melting temperature so the hybridized H and dissociated D states are equiprobable)
\item $G_{F4}$ = 5.1 KT
\item $f_H$ + $f_D$ + $f_F$ = 1 (the fractional populations sum to unity)
\item $f_X$ = $\exp (-\beta G_X$) / $Z$ ($Z$ is the partition function that imposes normalization)
\end{itemize}

We can solve this system of equations to yield $f_{F4}$ = 0.0030. A more rigorous accounting of the state populations by analyzing the MSM predictions yields a slightly lower prediction of $f_{F4}$ = 0.0028 that we report in Fig.~2b.

We recognize that this was a significant miscalculation, and we have re-checked all free energy results to ensure it was an error for the GC-core free energy calculations only and did not affect the reported values for GC-end and AT-all. Furthermore, we have verified that the error does not propagate to any other relevant results in the manuscript.

@@

We have updated Figure 3 with the correct $\Delta$ F  as shown below.

\begin{figure}[ht!]
	\begin{center} 
        \includegraphics[width=160mm, scale=1]{Fig3.pdf}
        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence melting temperatures}.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$. We define the hybridized state H to possess a free energy of zero and take care to only compare relative free energies (i.e., $\Delta F$) between the MSM and NN model. (b) Discrepancy between the macrostate relative free energy predictions $\Delta \Delta F = \Delta F^\mathrm{MSM} - \Delta F^\mathrm{NN}$ of the MSM relative to the NN model. The MSM tends to predict higher relative free energies (i.e., lower occupancy probabilities) for the out-of-register shifted states 5S2, 3S2, 5S4, 3S4 relative to the hybridized state H compared to the NN model.}
        \label{fig:NN_table}
	\end{center}
\end{figure}

We have also updated the second paragraph of section 3.3 as follows:

As illustrated in Fig.~3, we see that the MSM tends to predict higher free energies for \rood{all macrostates relative to the H state compared to the NN model}.

@@

\textcolor{blue}{10. I think some of the characterisation of the "all or nothing" picture is a bit misleading. The two state model of duplex formation is really a thermodynamic model, based on the assumption that the ensemble of hybridized and de-hybridized states look very different and are separated by a large barrier. This model is perfectly consistent with the transition between the two ensembles involving a complex network of metastable states. Indeed, as the authors note on line 3 of p4, a number of previous authors have considered what such states might look like, particularly for repetitive sequences. The authors' primary contribution here is not proposing that these states exist for the first time, but trying to tie together evidence from simulation and experiment to describe them.}

\textbf{Author Reply:}   We thank the reviewer for this important observation and illuminating our need to better clarify this potentially ambiguous concept. We have modified the text in Section 3.6 on p.~29 to read:

\noteg{Added sentence starting with ``Indeed, the sequence...'' Not totally sure about this, but I don't think we need to remove all references to all-or-nothing}
``The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} \rood{This characterization is intended to reflect that the mechanistic observation that} GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~2c), indicating that association and dissociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM. \rood{Indeed, the sequence's propensity to fray in simulation and the role of termini fraying in dissociation experiments \citep{Sanstead2016} indicate that there remain deviations from ``all-or-nothing'' behavior which are not fully captured within the MSM lag time. (We note that ``all-or-nothing'' can also be used, somewhat ambiguously, to refer to a two-state thermodynamic model where the hybridized and dehybridized are separated by a large free energy barrier but that the transitions between them may proceed through a network of metastable states.)}''

We have also removed ``all-or-nothing'' from the experimental methods section as the fitting procedure is more appropriately described as simply ``two-state.''

### 2021-07-28 SI_marked_up.tex
updated figS7

>>>>

\rood{In applying the NN models to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a. Indeed, each macrostate is composed of an ensemble of partially frayed states with higher free energy contributions than the intact pattern. However, when averaging across the frayed contributions of all microstates in each macrostate, we have found that incorporating additional NN configurations changes 
$\Delta\Delta$ F values by $<$ 1 kJ/mol when compared to using a single microstate. Given the extent to which a single NN representation captures each macrostate (which is further demonstrated in Fig.~S7), we have elected to maintain a single microstate free energy estimate for each macrostate}

### 2021-07-28 manuscript_marked_up.tex
updated figS7

To computationally mimic the T-jump process in our 3SPN.2 simulations, we conducted 1 $\mu$s simulations of an initially hybridized DNA duplex over a range of temperatures and monitored its structural relaxation. We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and $k_d^\mathrm{fast}$ (\blauw{Fig.~S4}). First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm starting from an initial fully hybridized duplex. This cutoff was selected as the distance beyond \rood{which} the strands are effectively non-interacting and defines the dissociated state. \rood{We verified that our results were robust to choices of this cutoff over the range 0.8-2.5 nm.} We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. \rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to correct our predictions for $f_\mathrm{hybridized}$ bulk yields. We correct the values observed in our simulations -- $f_\mathrm{hybridized}$ = $f_\mathrm{hybridized}^\mathrm{sim}$ -- to the bulk corrected result -- $f$ = $f_\mathrm{hybridized}$  -- using the following relations:}

    \begin{equation}\label{}
    \phi = \frac{\mathrm{f_\mathrm{hybridized}^\mathrm{sim}}}{\mathrm{1 - f_\mathrm{hybridized}^\mathrm{sim}}}
    \end{equation}

>>>>

To computationally mimic the T-jump process in our 3SPN.2 simulations, we conducted 1 $\mu$s simulations of an initially hybridized DNA duplex over a range of temperatures and monitored its structural relaxation. We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and $k_d^\mathrm{fast}$ (\blauw{Fig.~S4}). First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm starting from an initial fully hybridized duplex. This cutoff was selected as the distance beyond \rood{which} the strands are effectively non-interacting and defines the dissociated state. \rood{We verified that our results were robust to choices of this cutoff over the range 0.8-2.5 nm.} We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. \rood{To correct for finite-size effects associated with the simulation of a single duplex within a periodic simulation box and ensure that we will observe asymptotically correct behavior at long timescales, we apply the fluctuation correction proposed by Ouldridge et al.\citep{Ouldridge2010ExtractingSimulations} to our predictions for $f_\mathrm{hybridized}$ bulk yields. We correct the values observed in our simulations $f_\mathrm{hybridized}^\mathrm{sim}$ to the bulk corrected result $f_\mathrm{hybridized}$ using the following relations:}

    \begin{equation}\label{}
    \phi = \frac{f_\mathrm{hybridized}^\mathrm{sim}}{1 - f_\mathrm{hybridized}^\mathrm{sim}}
    \end{equation}
